---
title: "Profiling"
output: rmarkdown::html_vignette
bibliography: ref.bib 
description: >
  "Profiling" provides an overview of comprehensive analyses to efficiently 
   examine data quality, the clustering of samples, the correlation between 
  lipid species, and the composition of lipid characteristics.
vignette: >
  %\VignetteIndexEntry{Profiling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r load_package, message = FALSE, echo=FALSE}
library(dplyr)
library(utils)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(fig.width = 6, fig.height = 5)
```

"Profiling" provides an overview of comprehensive analyses to efficiently 
examine data quality, the clustering of samples, the correlation between 
lipid species, and the composition of lipid characteristics.

All of the input data of functions must be a SummarizedExperiment object constructed 
by `LipidSigR::as_summarized_experiment`. For detailed instructions for constructing 
SummarizedExperiment object, please read `vignette("1_tool_function")`

* NOTE: Some functions will require `processed_se`, which is the SummarizedExperiment 
object after being processed by `LipidSigR::data_process`. Please read `vignette("1_tool_function")`

```{r load LipidSigR, message = FALSE}
library(LipidSigR)
```

```{r load_data_DE, message = FALSE, include=FALSE, echo=FALSE}
data("de_data_twoGroup")
se <- de_data_twoGroup
# data processing
processed_se <- data_process(
    se, exclude_missing=TRUE, exclude_missing_pct=70, 
    replace_na_method='min', replace_na_method_ref=0.5, 
    normalization='Percentage')
```

# Cross-sample variability
There are three types of distribution plots, which could provide a simple view of 
sample variability a simple view of sample variability to compare the 
amount/abundance difference of lipid between samples (i.e., patients vs. control).

```{r Profiling: cross-sample variability, message = FALSE}
# conduct profiling
result <- cross_sample_variability(se)

# result summary
summary(result)
```

After running the above code, you will obtain a list called `result`, containing 
interactive plots, static plots, and tables for three types of distribution plots.
*(Note: Only static plots are displayed here.)*

```{r , fig.height=5, fig.width=8}
# view result: histogram of lipid numbers
result$static_lipid_number_barPlot
```

**Histogram of lipid numbers** 
The histogram overviews the total number of lipid species over samples. 
From the plot, we can discover the number of lipid species present in each sample.


```{r , fig.height=4.5, fig.width=8.2}
# view result: histogram of the total amount of lipid in each sample.
result$static_lipid_amount_barPlot  
```

**Histogram of lipid amount**
The histogram describes the variability of the total lipid amount between samples.


```{r , fig.width=7.7}
# view result: density plot of the underlying probability distribution
result$static_lipid_distribution
```

**Density plot of abundance distribution**
The density plot uncovers the distribution of lipid abundance in each sample (line). 
From this plot, we can have a deeper view of the distribution between samples.


# Dimensionality reduction
Dimensionality reduction is commonly used when dealing with large numbers of 
observations and/or large numbers of variables in lipids analysis. It transforms 
data from a high-dimensional space into a low-dimensional space so that it 
retains vital properties of the original data and is close to its intrinsic dimension.

Here we provide 3 dimensionality reduction methods, PCA, t-SNE, UMAP. 
As for the number of groups shown on the PCA, t-SNE, and UMAP plot, 
it can be defined by users (default: 2 groups). 

##  PCA
PCA (Principal component analysis) is an unsupervised linear dimensionality 
reduction and data visualization technique for high dimensional data, which tries 
to preserve the global structure of the data. Scaling (by default) indicates that 
the variables should be scaled to have unit variance before the analysis takes 
place, which removes the bias towards high variances. In general, scaling 
(standardization) is advisable for data transformation when the variables in the 
original dataset have been measured on a significantly different scale. As for 
the centering options (by default), we offer the option of mean-centering, 
subtracting the mean of each variable from the values, making the mean of each 
variable equal to zero. It can help users to avoid the interference of misleading 
information given by the overall mean.
```{r Profiling: dimensionality reduction - PCA}
# conduct PCA
result_pca <- dr_pca(
    processed_se, scaling=TRUE, centering=TRUE, clustering='kmeans', 
    cluster_num=2, kmedoids_metric=NULL, distfun=NULL, hclustfun=NULL, 
    eps=NULL, minPts=NULL, feature_contrib_pc=c(1,2), plot_topN=10)

# result summary
summary(result_pca)
```
After running the above code, you will obtain a list containing interactive plots, 
static plots, and tables for three types of distribution plots.
*(Note: Only static plots are displayed here.)*

```{r }
# view result: PCA plot
result_pca$static_pca
```

**PCA plot**

```{r}
# view result: scree plot of top 10 principle components
result_pca$static_screePlot
```

**Scree plot**
A common method for determining the number of PCs to be retained. The 'elbow' of the graph indicates all components to the left of this point can explain most variability of the samples

```{r}
# view result: correlation circle plot of PCA variables
result_pca$static_feature_contribution
```

**Correlation circle plot**
The correlation circle plot showing the correlation between a feature (lipid species) and a principal component (PC) used as the coordinates of the variable on the PC [@abdi2010principal]. The positively correlated variables are in the same quadrants while negatively correlated variables are on the opposite sides of the plot origin. The closer a variable to the edge of the circle, the better it represents on the factor map.


```{r}
# view result: bar plot of contribution of top 10 features
result_pca$static_variablePlot
```

**Bar plot of contribution of top 10 features**
The plot displaysthe features (lipid species) that contribute more to the user-defined principal component.


## t-SNE
t-SNE (t-Distributed Stochastic Neighbour Embedding) is an unsupervised 
non-linear dimensionality reduction technique that tries to retain the local 
structure(cluster) of data when visualising the high-dimensional datasets. 
Package `Rtsne` is used for calculation, and PCA is applied as a 
pre-processing step. In t-SNE, `perplexity` and `max_iter` are adjustable for 
users. The `perplexity` may be considered as a knob that sets the number of 
effective nearest neighbours, while `max_iter` is the maximum number of iterations 
to perform. The typical perplexity range between 5 and 50, but if the t-SNE 
plot shows a ‘ball’ with uniformly distributed points, you may need to lower 
your perplexity [@van2008visualizing].

```{r Profiling: dimensionality reduction - t-SNE}
# conduct t-SNE
result_tsne <- dr_tsne(
    processed_se, pca=TRUE, perplexity=5, max_iter=500, clustering='kmeans',
    cluster_num=2, kmedoids_metric=NULL, distfun=NULL, hclustfun=NULL, 
    eps=NULL, minPts=NULL)

# result summary
summary(result_tsne)
```

```{r }
# view result: t-SNE plot
result_tsne$static_tsne
```

**t-SNE plot**


##  UMAP 
UMAP (Uniform Manifold Approximation and Projection) using a nonlinear 
dimensionality reduction method, Manifold learning, which effectively visualizing 
clusters or groups of data points and their relative proximities. Both tSNE and 
UMAP are intended to predominantly preserve the local structure that is to group 
neighbouring data points which certainly delivers a very informative visualization 
of heterogeneity in the data. The significant difference with t-SNE is scalability, 
which allows UMAP eliminating the need for applying pre-processing step (such as PCA). 
Besides, UMAP applies Graph Laplacian for its initialization as tSNE by default 
implements random initialization. Thus, some people suggest that the key problem 
of tSNE is the Kullback-Leibler (KL) divergence, which makes UMAP superior over 
t-SNE. Nevertheless, UMAP’s cluster may not good enough for multi-class pattern 
classification [@mcinnes2018umap].
    
The type of distance metric to find nearest neighbors the size of the local 
neighborhood (as for the number of neighboring sample points) are set by 
parameter `metric` and `n_neighbors`. Larger values lead to more global views 
of the manifold, while smaller values result in more local data being preserved. 
Generally,  values are set in the range of 2 to 100. (default: 15).

```{r Profiling: dimensionality reduction - UMAP}
# conduct UMAP
result_umap <- dr_umap(
    processed_se, n_neighbors=15, scaling=TRUE, umap_metric='euclidean',
    clustering='kmeans', cluster_num=2, kmedoids_metric=NULL,
    distfun=NULL, hclustfun=NULL, eps=NULL, minPts=NULL)

# result summary
summary(result_umap)
```

```{r }
# view result: UMAP plot
result_umap$static_umap
```

**UMAP plot**

# Correlation heatmap
The correlation heatmap illustrates the correlation between samples or lipid
species and also depicts the patterns in each group. The correlation is
calculated by the method defined by parameter `corr_method`, and the
correlation coefficient is then clustered depending on method defined by
parameter `distfun` and the distance defined by parameter
`hclustfun`. Users can choose to output the sample correlation or lipid correlation 
results by the parameter `type`.

Please note that if the number of lipids or samples is over 50, the names of
lipids/samples will not be shown on the heatmap.

Here, we use `type='sample'` as example. 
```{r Profiling: correlation heatmap}
# correlation calculation
result_heatmap <- heatmap_correlation(
    processed_se, char=NULL, transform='log10', correlation='pearson', 
    distfun='maximum', hclustfun='average', type='sample')

# result summary          
summary(result_heatmap)
```

```{r }
# view result: sample-sample heatmap
result_heatmap$static_heatmap
```

**Heatmap of sample to sample correlations**
Correlations between lipid species are colored from strong positive correlations (red) to no correlation (white).

# Lipid characteristics
You can take a view of lipid expression over specific lipid
characteristics. First, lipids are classified by characteristics selected from
the 'Lipid characteristics' table. Here, we select "class" as the selected lipid
characteristic. The results will be showed by two plots.

* You can obtain the selectable lipid characteristics for the `char` input 
using `LipidSigR::list_lipid_char`. Please read `vignette("1_tool_function")`.

Here, we use `class` as the `char` input for an example.
```{r Profiling: lipid characteristics}
# calculate lipid expression of selected characteristic
result_lipid <- lipid_profiling(processed_se, char="class")

# result summary
summary(result_lipid)
```

```{r , fig.width=8}
# view result: bar plot
result_lipid$static_char_barPlot
```

**Bar plot classified by selected characteristic**
The bar plot depicts the abundance level of each sample within each group (e.g., PE, PC) of selected characteristics (e.g., class).
 
```{r , fig.width=8}
# view result: stacked horizontal bar chart
result_lipid$static_lipid_composition    
```

**Lipid class composition**
The stacked horizontal bar chart illustrates the percentage of characteristics 
in each sample. The variability of percentage between samples can also be obtained from this plot.
