---
title: "Correlation"
output: rmarkdown::html_vignette
bibliography: ref.bib 
description: >
     "Correlation" provides functions such as correlation coefficients and linear 
     regression to analyze continuous clinical features correlating with lipid species 
     or characteristics.
vignette: >
  %\VignetteIndexEntry{7. Correlation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r load_package, message = FALSE, echo=FALSE}
library(dplyr)
library(utils)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(fig.width = 6, fig.height = 5)
```

"Correlation" provides functions such as correlation coefficients and linear 
regression to analyze continuous clinical features correlating with lipid species 
or characteristics.

All of the input data of functions must be a SummarizedExperiment object constructed 
by `LipidSigR::as_summarized_experiment`. For detailed instructions for constructing 
SummarizedExperiment object, please read `vignette("tool_function")`.

* NOTE: Some functions will require `processed_se`, which is the SummarizedExperiment 
object after being processed by `LipidSigR::data_process`. Please read `vignette("tool_function")`.

To use our data as an example, follow the steps below.
```{r load LipidSigR, message=FALSE}
# load package
library(LipidSigR)

# load the example SummarizedExperiment
data("corr_data")

# data processing
processed_se <- data_process(
    corr_data, exclude_missing=TRUE, exclude_missing_pct=70, 
    replace_na_method='min', replace_na_method_ref=0.5, 
    normalization='Percentage')
```

Correlation analysis can be performed on either **"lipid species"** or **"lipid characteristics"**.

* **Lipid Species Analysis**: Data is analyzed at the individual lipid species level.
* **Lipid Characteristics Analysis**: Data is aggregated by specific lipid characteristics. 
The abundance of all lipid species in the same categories of a selected characteristic 
is summed up for analysis. 
    
This section is designed for continuous clinical data. The `condition_col` parameter 
specifies the column name from the group information table that defines the clinical 
conditions for analysis. Please ensure the selected columns contain only numerical data 
(e.g., integers, floats).
*NOTE: The condition_col must include at least two clinical conditions.*

Two correlation analyses are accessible: **'Correlation Coefficient'** and **'Linear Regression'**. 
Heatmaps will be shown once the correlation analysis is completed, it depicts the pattern 
between lipid species/lipid characteristics and clinical features. 
    
The available clustering methods are as follows. 

* Distance measurement: Pearson, Spearman, Kendall, Euclidean, Maximum, Manhattan, Canberra, Binary, and Minkowski.
* Clustering method: median, average, single, complete, Ward.D, Ward.D2, WPGMA, and UPGMC.

# Correlation coefficient {#correlation}
The Correlation Coefficient gives a summary view of whether a relationship exists 
between clinical features and lipid species, how strong that relationship is, 
and whether the relationship is positive or negative. The parameters' 
`cor_coef_cutoff` and `p_cutoff` can decide the correlation coefficient and p-value 
cut-offs. The rule of thumb in medical research recommended by Mukaka for 
interpreting the size of a correlation coefficient is provided below [@mukaka2012guide].


Size of Correlation          | Interpretation                             
-----------------------------|--------------------------------------------                   
0.90 to 1.00 (-.90 to -1.00) | Very high positive (negative) correlation
0.70 to .90 (-.70 to -.90)   | High positive (negative) correlation 
0.50 to .70 (-.50 to -.70)   | Moderate positive (negative) correlation
0.30 to .50 (-.30 to -.50)   | Low positive (negative) correlation     
0.00 to .30 (.00 to -.30)    | negligible correlation


Before the analysis, we must determine the `condition_col` parameter, selected from 
the column names from the group information table. Follow the steps below to view 
all available options for `condition_col`.
```{r}
# extract group information table
group_info <- extract_summarized_experiment(processed_se)$group_info

# condition_col options (choose only columns with numeric values)
condition_col_option <- str(group_info[-1])
```


## Lipid species correlation
For lipid species correlation analysis, set the `type` parameter to `Sp`. 
Optionally, you can visualize the heatmap with side colors by specifying a 
characteristic in the `side_color_char` parameter (e.g., 'class'). You can view 
all available characteristics from the common list returned by `LipidSigR::list_lipid_char`. 
If you prefer not to display side colors, set `side_color_char` to NULL.

*NOTE: The `char` parameter must be set to NULL for lipid species correlation.*
```{r Correlation_correlation_Sp}
# compute correlation coefficient and visualize by heatmap
res_sp <- corr_cor_heatmap(
    processed_se, char=NULL, 
    condition_col=c("FEV1_FVC", "Emphysema", "Exacerbations"),
    side_color_char='class', correlation='pearson', significant='pval',
    p_cutoff=1, adjust_p_method='BH', cor_coef_cutoff=0, 
    distfun='spearman', hclustfun='average', heatmap_col='statistic', 
    transform='log10', type='Sp')

# result summary
summary(res_sp)
```

```{r}
# view result: heatmap of clinical features and lipid species
res_sp$static_heatmap
```

**Correlation coefficient for lipid species analysis** The heatmap shows only the 
variables that pass the defined cut-offs for the p-value and the correlation 
coefficient. The rows are clinical features, and the columns are the lipid species.

## Lipid characteristics correlation
For lipid characteristics analysis, set the type parameter to `Char` and specify 
a lipid characteristic in the char parameter (e.g., 'class'). Use `LipidSigR::list_lipid_char`
to view the available lipid characteristics. For more detailed information, 
please refer to `vignette("tool_function")`.

*NOTE: The `side_color_char` parameter must be set to NULL for lipid characteristics correlation.*
```{r Correlation_correlation_char}
# compute correlation coefficient and visualize by heatmap
res_char <- corr_cor_heatmap(
    processed_se, char="class", 
    condition_col=c("FEV1_FVC", "Emphysema", "Exacerbations"),
    side_color_char=NULL, correlation='pearson', significant='pval',
    p_cutoff=1, adjust_p_method='BH', cor_coef_cutoff=0, 
    distfun='spearman', hclustfun='average', heatmap_col='statistic', 
    transform='log10', type='Char')

# result summary
summary(res_char)
```

```{r}
# view result: heatmap of clinical features and lipid characteristics
res_char$static_heatmap
```

**Correlation coefficient for lipid characteristics analysis** The heatmap shows 
only the variables that pass the defined cut-offs for the p-value and the 
correlation coefficient. The rows are clinical features, and the columns are 
lipid characteristics.


# Linear regression {#lr}
Linear regression is a statistical method used to model the relationship between 
a dependent variable (e.g., lipid level) and one or more independent variables 
(e.g., clinical features). By analyzing these relationships, we can estimate the 
impact of specific clinical factors on lipid levels.

In multiple linear regression, we incorporate additional variables to account for 
potential confounding factors. To do this, we specify the `adjusted_col` parameter, 
which refers to the column names selected from the group information table. Once 
the analysis is complete, each lipid species is assigned a beta coefficient and 
a corresponding t-statistic (p-value). These values can be used to identify 
significant associations and group similar lipid species for further analysis.

Before the analysis, we must determine the `condition_col` and `adjusted_col` 
parameters, selected from the column names from the group information table. 
Follow the steps below to view all available options for `condition_col` and `adjusted_col`.
```{r}
# extract group information table
group_info <- extract_summarized_experiment(processed_se)$group_info

# condition_col options (choose only columns with numeric values)
condition_col_option <- str(group_info[-1])

# adjusted_col options
adjusted_col_option <- str(group_info[-1])
```


## Lipid species correlation
For lipid species correlation analysis, set the `type` parameter to `Sp`. 
Optionally, you can visualize the heatmap with side colors by specifying a 
characteristic in the `side_color_char` parameter (e.g., 'class'). You can view 
all available characteristics from the common list returned by `LipidSigR::list_lipid_char`. 
If you prefer not to display side colors, set `side_color_char` to NULL.

*NOTE: The `char` parameter must be set to NULL for lipid species correlation.*
```{r Correlation_lr_sp}
# compute linear regression and visualize by heatmap
res_sp <- corr_lr_heatmap(
    processed_se, char=NULL,
    condition_col=c("FEV1_FVC", "Emphysema", "Exacerbations"),
    adjusted_col=c("Age", "Sex", "Smoking", "BMI", "FEV1"),
    side_color_char=NULL, significant='pval', p_cutoff=0.05,
    adjust_p_method='BH', distfun='spearman', hclustfun='centroid',
    heatmap_col='t_statistic', transform='log10', type='Sp')

# result summary
summary(res_sp)
```

```{r}
# view result: heatmap of linear regression
res_sp$static_heatmap
```

**The heatmap of linear regression for lipid species analysis** The heatmap shows 
only the variables that pass the user-defined cut-offs for p-value and correlation 
coefficient. The rows are clinical features, and the columns are lipid species.

## Lipid characteristics correlation
For lipid characteristics correlation, set the `type` parameter to `Char` and 
specify a lipid characteristic for the `char` parameter. You can use 
`LipidSigR::list_lipid_char` to view a list of available lipid characteristics. 
For more detailed information, please refer to `vignette("tool_function")`.

*NOTE: The `side_color_char` parameter must be set to NULL for lipid characteristics correlation.*
```{r Correlation_lr_char}
# compute linear regression and visualize by heatmap
res_char <- corr_lr_heatmap(
    processed_se, char="class",
    condition_col=c("FEV1_FVC", "Emphysema", "Exacerbations"),
    adjusted_col=c("Age", "Sex", "Smoking", "BMI", "FEV1"),
    side_color_char=NULL, significant='pval', p_cutoff=1,
    adjust_p_method='BH', distfun='spearman', hclustfun='centroid',
    heatmap_col='t_statistic', transform='log10', type='Char')

# result summary
summary(res_char)
```

```{r}
# view result: heatmap of linear regression
res_char$static_heatmap
```

**The heatmap of linear regression for lipid characteristics analysis** Only the 
variables that pass the user-defined cut-offs for p-value and correlation coefficient 
are shown on the heatmap. The rows of the heatmap are clinical features, and the 
columns are lipid characteristics.
