---
title: "Differential expression"
output: rmarkdown::html_vignette
bibliography: ref.bib 
description: >
  "Differential expression" integrates many useful lipid-focused analyses for identifying 
    significant lipid species or lipid characteristics.
vignette: >
  %\VignetteIndexEntry{Differential expression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r load_package, message = FALSE, echo=FALSE}
library(dplyr)
library(utils)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(fig.width = 6, fig.height = 5)
```

"Differential expression" integrates many useful lipid-focused analyses for 
identifying significant lipid species or lipid characteristics.

Differential Expression is divided into two main analyses, **'Lipid species analysis'**
and **'Lipid characteristics analysis'**. Further analysis and
visualization methods can also be conducted based on the results of differential
expressed analysis.

* **Lipid species analysis**: The lipid species analysis explores the
  significant lipid species based on differentially expressed analysis. Data are
  analyzed based on each lipid species. Further analysis and visualization
  methods, include

  1. dimension reduction,
  2. hierarchical clustering,
  3. characteristics association.

* **Lipid characteristics analysis**: The lipid characteristics
  analysis explores the significant lipid characteristics. Lipid species are
  categorized and summarized into a new lipid abundance table according to a
  selected lipid characteristic. The abundance of all lipid species of the same
  categories are summed up, then conduct differential expressed analysis.
  Further analysis and visualization methods include

  1. dimension reduction, 
  2. hierarchical clustering.

All input data for differential expression and the downstream functions must 
be a SummarizedExperiment object constructed using `LipidSigR::as_summarized_experiment` 
and then further processed with` LipidSigR::data_process`. 
Please read `vignette("tool_function")` before preparing the input data.

To use our data as an example, follow the steps below.
```{r load LipidSigR, message=FALSE}
# load package
library(LipidSigR)

# load the example SummarizedExperiment
data("de_data_twoGroup")

# data processing
processed_se <- data_process(
    de_data_twoGroup, exclude_missing=TRUE, exclude_missing_pct=70, 
    replace_na_method='min', replace_na_method_ref=0.5, 
    normalization='Percentage', transform='log10')
```


# Lipid species differential expression analysis
For lipid species analysis section, differential expression analysis is
performed to figure out significant lipid species. In short, samples will be
divided into independent groups according to the input "Group Information" table.

Here, we use two-group data as an example.
```{r DE_lipid species: differentially expressed analysis}
# conduct differential expression analysis of lipid species
deSp_se <- deSp_twoGroup(
    processed_se, ref_group='ctrl', test='t-test',
    significant='pval', p_cutoff=0.05, FC_cutoff=1, transform='log10')
```

* NOTE: For multi-group data, please use `LipidSigR::deSp_multiGroup`.

After running the above code, a SummarizedExperiment object `deSp_se` will be 
returned containing the analysis results. This object can be used as input for 
plotting and further analyses such as [dimensionality reduction](#desp-dr), 
[hierarchical clustering](#desp-clustering), and
[characteristics association](#desp-char)

`deSp_se` includes the input abundance data, lipid characteristic table, 
group information table, analysis results, and input parameter settings. 
You can view the data in `deSp_se` by `LipidSigR::extract_summarized_experiment`.
Please read `vignette("tool_function")`.

The differential expression analysis result can be input for plotting MA plots, 
volcano plots, and lollipop plots. *(Note: Only static plots are displayed here.)*
```{r}
# plot differential expression analysis result
deSp_plot <- plot_deSp_twoGroup(deSp_se)

# result summary
summary(deSp_plot)
```

```{r}
# view result: lollipop chart
deSp_plot$static_de_lipid
```

**Lollipop chart of lipid species analysis**
The lollipop chart reveals the lipid species that pass chosen cut-offs. The x-axis shows log2 fold change while the y-axis is a list of lipids species. The color of the point is determined by `-log10(adj_value/p-value)`.

```{r }
# view result: MA plot
deSp_plot$static_maPlot
```

**MA plot**
The MA plot indicates three groups of lipid species, up-regulated(red), down-regulated(blue), and non-significant(grey).
 
```{r}
# view result: MA plot
deSp_plot$static_volcanoPlot 
```

**Volcano plot**
The volcano plot illustrates a similar concept to the MA plot. These points visually identify the most biologically significant lipid species (red for up-regulated, blue for down-regulated, and grey for non-significant).

You can further plot an abundance box plot for any lipid species of interest by 
`LipidSigR::boxPlot_feature_twoGroup`. (for multi-group data, please use `LipidSigR::boxPlot_feature_multiGroup`)

For example, let's use `TAG 48:0;0`, a significant lipid species from the lollipop above.
```{r boxPlot}
# plot abundance box plot of 'TAG 48:0;0'
boxPlot_result <- boxPlot_feature_twoGroup(
    processed_se, feature='TAG 48:0;0', ref_group='ctrl', test='t-test',
    transform='log10')

# result summary
summary(boxPlot_result)

# view result: static box plot
boxPlot_result$static_boxPlot
```

**Box plot of lipid abundance** An asterisk sign indicates significant differences between groups. 
The absence of an asterisk or line denotes a non-significant difference between groups.


# Lipid characteristics differential expression analysis {#dechar}
The massive degree of structural diversity of lipids contributes to the
functional variety of lipids. The characteristics can range from subtle
variance (i.e. the number of a double bond in the fatty acid) to major change
(i.e. diverse backbones). In this section, lipid species are categorized and
summarized into a new lipid abundance table according to two selected lipid
characteristics, then conducted differential expressed analysis.

Two procedures of analysis will be conducted - first is **'Characteristics'** 
and then **'Subgroup of characteristics'**. 

**'Characteristics'** is based on the first selected 'characteristics' 
while **'Subgroup of characteristics'** is the subgroup analysis of the previous section. 
Analyses will be performed based on parameter `char` and `subChar` selected by users.

Before beginning, we suggest calculating the two-way ANOVA and reviewing the results for all lipid characteristics.
```{r two_way_anova}
# two way anova
twoWayAnova_table <- char_2wayAnova(
    processed_se, ratio_transform='log2', char_transform='log10')

# view result table
head(twoWayAnova_table[, 1:4], 5) 
```

From the table returned by `LipidSigR::char_2wayAnova`, we have to selected the 
lipid characteristics of interest as `char` and `subChar` 
for characteristics and subgroup of characteristics analyses.

Here, we use two-group data as an example, and choose `Total.C` as the `char` input.
```{r DE_lipid characteristics: differentially expressed analysis}
# conduct differential expression of lipid characteristics
deChar_se <- deChar_twoGroup(
    processed_se, char="Total.C", ref_group="ctrl", test='t-test', 
    significant="pval", p_cutoff=0.05, FC_cutoff=1, transform='log10')
```

* NOTE: For multi-group data, please use `LipidSigR::deChar_multiGroup`.

After running the above code, a SummarizedExperiment object `deChar_se` will be 
returned containing the analysis results. This object can be used as input for 
plotting and further analyses such as [dimension reduction](#dechar-dr), and
[hierarchical clustering](dechar-clustering).

`deChar_se` includes the input abundance data, lipid characteristic table, 
group information table, analysis results, and input parameter settings. 
You can view the data in `deChar_se` by `LipidSigR::extract_summarized_experiment`. 

Next, you can plot the differential expression analysis result by `LipidSigR::plot_deChar_twoGroup`. 
For multi-group data, please use `LipidSigR::plot_deChar_multiGroup`.
```{r }
# plot differential expression analysis results
deChar_plot <- plot_deChar_twoGroup(deChar_se)

# result summary
summary(deChar_plot)
```

**The results of 'Characteristics' analysis in the first section**

```{r }
# view result: bar plot of selected `char`
deChar_plot$static_barPlot
# view result: sqrt-scaled bar plot of selected `char`
deChar_plot$static_barPlot_sqrt
# view result: line plot of `selected char`
deChar_plot$static_linePlot
# view result: sqrt-scaled line plot of selected `char`
deChar_plot$static_linePlot_sqrt  
# view result: box plot of selected `char`
deChar_plot$static_boxPlot  
```

In the **'Subgroup of characteristics'**, besides the selected characteristic in 
first section defined by parameter `char`, we can further choose another 
characteristic by parameter `subChar`. The two chosen characteristics, 
`char` and `subChar`should be either both continuous data or one continuous and 
one categorical data.

* NOTE: You can use `LipidSigR::list_lipid_char` to get all the selectable lipid characteristics. 
Please read `vignette("tool_function")`.

Here we use `Total.C` as the `char` input and `class` as the `subChar` input for an example.
```{r DE_lipid characteristics: differentially expressed analysis_Subgroup}
# subgroup differential expression of lipid characteristics
subChar_se <- subChar_twoGroup(
    processed_se, char="Total.C", subChar="class", ref_group="ctrl", 
    test='t-test', significant="pval", p_cutoff=0.05, 
    FC_cutoff=1, transform='log10')
```

* NOTE: For multi-group data, please use `LipidSigR::subChar_multiGroup`.

After running the code, the returned `subChar_se` contained the input abundance 
data, lipid characteristic table, group information table, analysis results, 
and input parameter settings. You can view the data in `subChar_se` by `LipidSigR::extract_summarized_experiment`. Please read `vignette("tool_function")`.

Next, you can plot the results of a specific feature within the `subChar`. 
As we have chosen `class` as `subChar` in conducting `LipidSigR::subChar_twoGroup`, 
we can choose a feature within the `class` by parameter `subChar_feature` for plotting result plots.
Obtain all the selectable features for `subChar_feature` by following:
```{r}
# get subChar_feature list
subChar_feature_list <- unique(
    extract_summarized_experiment(subChar_se)$all_deChar_result$sub_feature)
```

Here we choose `Cer` as the `subChar_feature` input for an example.
For multi-group data, please use `LipidSigR::plot_subChar_multiGroup`.
``` {r DE_lipid characteristics: differentially expressed analysis_Subgroup plot}
# visualize subgroup differential expression of lipid characteristics
subChar_plot <- plot_subChar_twoGroup(subChar_se, subChar_feature="Cer")

# result summary
summary(subChar_plot)
```

*(Note: Only static plots are displayed here.)*

**The results of 'Subgroup of characteristics' analysis in the second section**

* *Note: The star above the bar shows the significant difference of the specific subgroup of the selected characteristic between control and experimental groups.*
```{r }
# view result: bar plot of `subChar_feature`
subChar_plot$static_barPlot  
# view result: sqrt-scaled bar plot of `subChar_feature`
subChar_plot$static_barPlot_sqrt 
# view result: line plot of `subChar_feature`
subChar_plot$static_linePlot 
# view result: sqrt-scaled line plot of `subChar_feature`
subChar_plot$static_linePlot_sqrt 
# view result: box plot of `subChar_feature`
subChar_plot$static_boxPlot  
```


# Dimension reduction {#desp-dr}
Four dimension reduction methods can be applied after lipid species or lipid 
characteristic analysis. Three methods—PCA, t-SNE, and UMAP—have already been introduced. 
Please read `Dimension reduction` in `vignette("profiling")`

* Note: The input data of this section must be the output `deSp_se` from [lipid species analysis](#desp),
or `deChar_se` from [lipid characteristics analysis](#dechar).

## PLS-DA 
```{r DE_lipid species: dimensionality reduction - PLS-DA}
# conduct PLSDA
result_plsda <- dr_plsda(
  deSp_se, ncomp=2, scaling=TRUE, clustering='group_info', cluster_num=2, 
  kmedoids_metric=NULL, distfun=NULL, hclustfun=NULL, eps=NULL, minPts=NULL)

# result summary
summary(result_plsda)
```

```{r }
# view result: PLS-DA plot
result_plsda$static_plsda
```

**PLS-DA plot**

```{r}
# view result: PLS-DA loading plot
result_plsda$static_loadingPlot    
```

**Loading plot**
In the PLS-DA loading plot, the distance to the center of the variables indicates the contribution of the variable. The value of the x-axis reveals the contribution of the variable to PLS-DA-1, whereas the value of the y-axis discloses the contribution of the variable to PLS-DA-2.

# Hierarchical clustering {#desp-clustering}
Based on the results of differential expression analysis, we further take a
look at differences of lipid species between the control group and the
experimental group. Lipid species derived from two groups are clustered and
visualized on heatmap by hierarchical clustering. Users can choose to output the
results of all lipid species or only significant lipid species by the parameter `type`.
    
The top of the heatmap is grouped by sample group (top annotation) while the side 
of the heatmap (row annotation) can be chosen from `lipid_char_table`, such as class, 
structural category, functional category, total length, total double bond (Total.DB), 
hydroxyl group number (Total.OH), the double bond of fatty acid(FA.DB), 
hydroxyl group number of fatty acid(FA.OH).

* Note: The input data of this section must be the output `deSp_se` from [lipid species analysis](#desp),
or `deChar_se` from [lipid characteristics analysis](#dechar).

* You can obtain the selectable lipid characteristics for the `char` input using `LipidSigR::list_lipid_char`. 
Please read `vignette("tool_function")`.

Here, we use `class` as the `char` input for an example.
```{r DE_lipid species: hierarchical clustering}
# conduct hierarchical clustering
result_hcluster <- heatmap_clustering(
    de_se=deSp_se, char='class', distfun='pearson', 
    hclustfun='complete', type='sig')

# result summary
summary(result_hcluster)
```

```{r}
# view result: heatmap of significant lipid species
result_hcluster$static_heatmap
```

**Heatmap of significant lipid species**
The difference between the two groups by observing the distribution of lipid species.


# Characteristics analysis {#desp-char}
The characteristics analysis visualizes the difference between control and 
experimental groups of significant lipid species categorized based on different 
lipid characteristics from `lipid_char_table`, such as class, 
structural category, functional category, total length, total double bond (Total.DB), 
hydroxyl group number (Total.OH), the double bond of fatty acid(FA.DB), 
hydroxyl group number of fatty acid(FA.OH).

* Note: The input data of this section must be the output `deSp_se` from [lipid species analysis](#desp).

* You can obtain the selectable lipid characteristics for the `char` input 
using `LipidSigR::list_lipid_char`. Please read `vignette("tool_function")`.

Here, we use `class` as the `char` input for an example.
```{r DE_lipid species: characteristics analysis}
# conduct characteristic analysis
result_char <- char_association(deSp_se, char='class')

# result summary
summary(result_char)
```

```{r }
# view result: bar chart
result_char$static_barPlot
```

**The bar chart of significant groups**
The bar chart shows the significant groups (values) with mean fold change over 2 in the selected characteristics by colors (red for significant and black for insignificant).

```{r }
# view result: lollipop plot
result_char$static_lollipop
```

**The lollipop chart of all significant groups**
The lollipop chart compares multiple values simultaneously and it aligns the log2(fold change) of all significant groups (values) within the selected characteristics.

```{r , fig.small=TRUE, fig.width=3, fig.height=3}
# view result: word cloud 
result_char$static_wordCloud
```

**Word cloud with the count of each group**
The word cloud shows the count of each group(value) of the selected characteristics.


# Double bond-chain length analysis
This section provides heatmaps that illustrates the correlation between the 
double bond and chain length of lipid species. The color in the heatmaps is 
gradient according to log2FC.

The correlation is visually represented by cell colors—red indicates a 
positive correlation, while blue indicates a negative. 
Significant correlations are highlighted with an asterisk sign on the plot.

* You can obtain the selectable lipid characteristics for the `char` input 
using `LipidSigR::list_lipid_char`. Please read `vignette("tool_function")`.

Here, we use `class` as the `char` input for an example.
```{r DE_lipid characteristics: two-heatmap all, fig.width=7.8}
# conduct double bond-chain length analysis (without setting `char_feature`)
heatmap_all <- heatmap_chain_db(
    processed_se, char='class', char_feature=NULL, ref_group='ctrl', 
    test='t-test', significant='pval', p_cutoff=0.05, 
    FC_cutoff=1, transform='log10')

# result summary 
summary(heatmap_all)

# summary of total chain result
summary(heatmap_all$total_chain)

# view result: heatmap of total chain
heatmap_all$total_chain$static_heatmap
# summary of each chain result
summary(heatmap_all$each_chain)

# view result: heatmap of each chain
heatmap_all$each_chain$static_heatmap
```

```{r DE_lipid characteristics: two-heatmap one}
# conduct double bond-chain length analysis (a specific `char_feature`)
heatmap_one <- heatmap_chain_db(
    processed_se, char='class', char_feature='PC', ref_group='ctrl',
    test='t-test', significant='pval', p_cutoff=0.05, 
    FC_cutoff=1, transform='log10')

# result summary 
summary(heatmap_one)

# summary of total chain result
summary(heatmap_one$total_chain)

# view result: heatmap of total chain
heatmap_one$total_chain$static_heatmap
# summary of each chain result
summary(heatmap_one$each_chain)

# view result: heatmap of each chain
heatmap_one$each_chain$static_heatmap
```

The data in `chain_db_se` can be viewed using `LipidSigR::extract_summarized_experiment`. 
Please read `vignette("tool_function")`.

You can further plot an abundance box plot for any lipid species of interest by 
`LipidSigR::boxPlot_feature_twoGroup`. For multi-group data, please use `LipidSigR::boxPlot_feature_multiGroup`.

For example, let's use `15:0`, a significant lipid species from the heatmap above.
```{r boxPlot_two_heatmap}
# plot abundance box plot of "15:0"
boxPlot_result <- boxPlot_feature_twoGroup(
    heatmap_one$each_chain$chain_db_se, feature='15:0', 
    ref_group='ctrl', test='t-test', transform='log10')

# result summary
summary(boxPlot_result)

# view result: static box plot
boxPlot_result$static_boxPlot
```

**Box plot of lipid abundance** An asterisk sign indicates significant differences between groups. 
The absence of an asterisk or line denotes a non-significant difference between groups.

# Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
