---
title: "LipidSigR tutorial"
author:
- name: Wei-Chung Cheng
  affiliation: China Medical University
output:
  #BiocStyle::pdf_document: default
  BiocStyle::html_document:
    toc_float: true
bibliography: ref.bib
csl: nature.csl
abstract: >
   "LipidSigR" is an R package developed based on LipidSig web-based tool (http://www.chenglab.cmu.edu.tw/lipidsig/). This package integrates a comprehensive analysis for streamlined data mining of lipidomic datasets. We provide 4 main analysis workflows, which is "Profiling", "Differential expression", "Machine learning", and "Correlation". Each section provides unique aspects to analyze the lipidome profiling data based on different characteristics including lipid class, chain length, unsaturation, hydroxyl group, and fatty acid composition.
vignette: |
  %\VignetteIndexEntry{LipidSigR tutorial}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

`r R.version.string`

Bioconductor version `r BiocManager::version()`

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(fig.width = 6, fig.height = 5)
knitr::opts_chunk$set(fig.align ="center")
```

```{r style, echo = FALSE, results = 'asis'} 
        BiocStyle::markdown()
```

# Introduction
Lipidomics technology provides a fast and high-throughput screening to identify thousands of lipid species in cells, tissues, or other biological samples and has been broadly used in several areas of studies. 

**"LipidSigR"** is an R package developed based on LipidSig web-based tool [http://www.chenglab.cmu.edu.tw/lipidsig/](http://www.chenglab.cmu.edu.tw/lipidsig/) [@lipidsig]. This package integrates a comprehensive analysis for streamlined data mining of lipidomic datasets. We provide 4 main analysis workflows, which is **"Profiling"**, **"Differential expression"**, **"Machine learning"**, and **"Correlation"**. Each section provides unique aspects to analyze the lipidome profiling data based on different characteristics including lipid class, chain length, unsaturation, hydroxyl group, and fatty acid composition.

The usage of the 4 sections is introduced briefly as below. 

* **Profiling**: The profiling section provides an overview of comprehensive analyses for you to efficiently examine data quality, the clustering of samples, the correlation between lipid species, and the composition of lipid characteristics.
* **Differential expression**: The differential expression section integrates many useful lipid-focused analyses for identifying significant lipid species or lipid characteristics.
* **Machine learning**: The machine learning section provides a broad variety of feature selection methods and classifiers to build binary classification models. Furthermore, the subsequent analyses can help users to evaluate the learning algorithm's performance and to explore important lipid-related variables. 
* **Correlation**: The correlation section illustrates and compares the relationships between different clinical phenotypes and lipid features. 

The functions and usage of the above 4 sections are orderly described in the following sections, which is Profiling in Section \@ref(Profiling), Differential expression in Section \@ref(DE), Machine learning in Section \@ref(ML), and Correlation in Section \@ref(corr).

# Installation 
Here is the procedures of running the **LipidSigR** package on your system. We assume that you have already installed the R program (see the R project at [http://www.r-project.org](http://www.r-project.org)  and are familiar with it. You need to have R 4.0.0 or a later version installed for running LipidSigR.

Our package is available at the Bioconductor repository [http://www.bioconductor.org](http://www.bioconductor.org). To install our package, first, you need to install the core Bioconductor packages. If you have already installed the Bioconductor packages on your system, you can skip the following step below.

```{r install_Bioconductor, eval=FALSE}
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install()
```

Once the core Bioconductor packages have been installed, we can begin the installation of the LipidSigR package.

```{r install_package, eval=FALSE}
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install("LipidSigR")
``` 

After conducting the above step, now you can load in our package and start using it!

```{r load_package, message = FALSE}
library(LipidSigR)
```
```{r , message = FALSE, echo=FALSE}
library(plotly)
library(pathview)
library(SHAPforxgboost)
library(visNetwork)
```


# Session info
```{r sessionInfo, echo=FALSE}
sessionInfo()
```

# References

