---
title: "Tool function"
output: rmarkdown::html_vignette
bibliography: ref.bib 
description: >
  The tool function for LipidSigR
vignette: >
  %\VignetteIndexEntry{1. Tool function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r load_package, message = FALSE, echo=FALSE}
library(dplyr)
library(utils)
library(rgoslin)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(fig.width = 6, fig.height = 5)
```

LipidSigR offers several utility functions to enhance convenience in constructing 
input SummarizedExperiment object, viewing output results, listing selectable 
lipid characteristics, and more.

```{r load LipidSigR, message = FALSE}
library(LipidSigR)
```

# Construct SE object
The input data for most functions must be a SummarizedExperiment object, 
constructed by `LipidSigR::as_summarized_experiment` or generated from other 
upstream analysis functions.

To begin analyzing data, follow the instructions below to construct the input 
SummarizedExperiment object. First, prepare the required input data frames: the 
abundance data and the group information table. If you only intend to conduct 
the profiling analysis, you only need to prepare the abundance data.

## Prepare input data frames
The input abundance data and group information table must be provided as data 
frames and adhere to the following requirements.

**Abundance data**: The lipid abundance data includes the abundance values 
of each feature across all samples. 

1. The first column of abundance data must contain a list of lipid names (features). 
1. Each lipid name (feature) is unique.
1. All abundance values are numeric.

For example:
```{r ex_abundance}
rm(list = ls())
data("abundance_twoGroup")
head(abundance_twoGroup[, 1:6], 5)
```

**Group information table**: The group information table contains the grouping 
details corresponding to the samples in lipid abundance data.

1. For two-group data, column names must be arranged in order of `sample_name`, 
`label_name`, `group`, and `pair`, and for multi-group data, column names must be arranged 
in order of `sample_name`, `label_name`, and `group`.
1. All sample names are unique.
1. Sample names in `sample_name` column are as same as the sample names in lipid abundance data.
1. Columns of `sample_name`, `label_name`, and `group` columns do not contain NA values.
1. For two-group data, the column `group` contain 2 groups, and for multi-group 
data the column `group` must contain more than 2 groups.
1. In the 'pair' column for paired data, each pair must be sequentially numbered 
from 1 to N, ensuring no missing, blank, or skipped numbers are missing; 
otherwise, the value should be all marked as NA. 
*(NOTE: The group information table of multi-group data should not contain this column.)*

For example:
```{r ex_groupInfo}
data("group_info_twoGroup")
head(group_info_twoGroup, 5)
```

## Mapping lipid characteristics {#lipid-map}
The purpose of this step is to exclude lipid features not recognized 
by `rgoslin::parseLipidNames`. 

If you haven't install `rgoslin` package, install it by running the following codes.
```{r install rgoslin, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("rgoslin")
```

Then, follow the instructions below before 
constructing the input data as a SummarizedExperiment object.

* In this step, an error message will be returned if `rgoslin::parseLipidNames` cannot 
recognize a certain lipid. However, if your data contains at least two recognizable lipids, 
it will be sufficient for analysis *(note that different analyses may have varying data requirements)*.

```{r rgoslin_twoGroup}
library(dplyr)
library(rgoslin)
# map lipid characteristics by rgoslin
parse_lipid <- rgoslin::parseLipidNames(lipidNames=abundance_twoGroup$feature)

# filter lipid recognized by rgoslin
recognized_lipid <- parse_lipid$Original.Name[
    which(parse_lipid$Grammar != 'NOT_PARSEABLE')]
abundance <- abundance_twoGroup %>% 
    dplyr::filter(feature %in% recognized_lipid)
goslin_annotation <- parse_lipid %>% 
    dplyr::filter(Original.Name %in% recognized_lipid)
```

After running the above code, two data frames, `abundance`, and 
`goslin_annotation`, will be generated and used in the next step.
```{r rgoslin_output}
head(abundance[, 1:6], 5)
head(goslin_annotation[, 1:6], 5)
```

## Construct SE object
Use the data obtained from previous steps to construct SE object.
```{r as_summarized_experiment}
se <- as_summarized_experiment(
    abundance, goslin_annotation, group_info=group_info_twoGroup, 
    se_type='de_two', paired_sample=FALSE)
```
After running the above code, you are ready to begin the analysis with the 
output `se`. After the code execution, a summary of the input data will be 
displayed. 

*(Note: If errors occur during execution, please revise the input data to resolve them.)*

# Extract data in SE object
Most of our statistical functions return results in a SummarizedExperiment object. 
To enhance user accessibility, we provide a function to extract these results as 
several data frames for easier viewing.

For example, after conducting `LipidSigR::deSp_twoGroup`, it returns `deSp_se`. 
You can view the data stored in the returned SummarizedExperiment object 
using `LipidSigR::extract_summarized_experiment`.
```{r conduct deSp}
# data processing
processed_se <- data_process(
    se, exclude_missing=TRUE, exclude_missing_pct=70, 
    replace_na_method='min', replace_na_method_ref=0.5, 
    normalization='Percentage', transform='log10')
# conduct differential expression analysis of lipid species
deSp_se <- deSp_twoGroup(
    processed_se, ref_group='ctrl', test='t-test',
    significant='pval', p_cutoff=0.05, FC_cutoff=1, transform='log10')
```
```{r extract_se}
# extract results in SE
res_list <- extract_summarized_experiment(deSp_se)
# summary of extract results
summary(res_list)
```

The returned result list, extracted from the SummarizedExperiment object, 
includes input data (abundance, lipid characteristics table, group information 
table), various input settings (e.g., significance level, p-value cutoff), 
and statistical results.


# Data processsing {#data-processing}
Most input data for `LipidSigR` must undergo data processing to ensure it is 
normalized without missing values. The data processing function operates on the constructed 
SummarizedExperiment object, processing the abundance table based on the user's settings. The 
resulting SummarizedExperiment object can then be used directly in subsequent analysis functions.

```{r data_processing}
# abundance in input SE
head(extract_summarized_experiment(se)$abundance[, 1:5], 10)

# data processing
processed_se <- data_process(
    se, exclude_missing=TRUE, exclude_missing_pct=70,
    replace_na_method='min', replace_na_method_ref=0.5,
    normalization='Percentage', transform='log10')

# abundance in processed SE
head(extract_summarized_experiment(processed_se)$abundance[, 1:5], 10)
```

After data processing, you can further visualize the differences by plotting the 
abundance before and after processing.
```{r data_processing_plot}
# plotting
data_process_plots <- plot_data_process(se, processed_se)

# result summary
summary(data_process_plots)

# view box plot before/after data processing
data_process_plots$static_boxPlot_before
data_process_plots$static_boxPlot_after

# view density plot before/after data processing
data_process_plots$static_densityPlot_before  
data_process_plots$static_densityPlot_after  
```


# Obtain lipid characteristics
In several functions, you must select a specific lipid characteristic as input 
for analysis. To enhance accessibility, we provide the `LipidSigR::list_lipid_char` 
function, which returns all available lipid characteristics. You can review these 
options and choose one as your input.

`LipidSigR::list_lipid_char` returns three types of lipid characteristic 
lists: `deChar_list`, `chain_db_list`, and `common_list`, each used in different analyses.

1. deChar_list: the selectable lipid characteristics for using `LipidSigR::deChar_twoGroup` and `LipidSigR::deChar_multiGroup`.

1. chain_db_list: the selectable lipid characteristics for using `LipidSigR::heatmap_chain_db`.

1. ml_char_list: the selectable lipid characteristics for using `LipidSigR::ml_model`.

1. common_list: the selectable lipid characteristics for use in all functions not mentioned above.

Take `LipidSigR::lipid_profiling` as an example.
```{r list_char ex}
# data processing
processed_se <- data_process(
    se, exclude_missing=TRUE, exclude_missing_pct=70,
    replace_na_method='min', replace_na_method_ref=0.5,
    normalization='Percentage', transform='log10')

# get lipid characteristics
list_lipid_char(processed_se)$common_list

# conduct lipid profiling function
result <- lipid_profiling(processed_se, char="class")
```


# Lipid annotation
"Mapping lipid characteristics" is one of the steps in constructing the input 
SummarizedExperiment object in `LipidSigR`, detailed in [previous section](#lipid-map).

When the Goslin annotation table is provided to the `LipidSigR::as_summarized_experiment`, 
it is extended into a lipid characteristic table by adding mappings between lipids, 
the LION ontology, and other resource IDs. This extended lipid characteristic table 
is then included in the returned SE object. If you only need the lipid characteristic table, 
you can use this `LipidSigR::lipid_annotation` directly.

```{r lipid_annotation}
# the input lipid annotation table
head(goslin_annotation[, 1:6], 5)

# conduct lipid annotation
lipid_annotation_table <- lipid_annotation(goslin_annotation)

# view lipid annotation table
head(lipid_annotation_table[, 1:5], 5)

# columns of returned lipid annotation table
colnames(lipid_annotation_table)
```


# Lipid species abundance conversion
Two conversion types for lipid species are provided: summing lipid species 
abundance by lipid characteristics or converting to ratio-based abundance.

Both functions are integrated into the lipid characteristics analysis function, 
so when you use `LipidSigR::deChar_twoGroup` or `LipidSigR::deChar_multiGroup`, 
the abundance conversion is automatically performed within the function.

If you only want to obtain the converted abundance, follow the commands below.

* NOTE: The input data must be a SummarizedExperiment object constructed using 
`LipidSigR::as_summarized_experiment` and then further processed with 
` LipidSigR::data_process`. 
Please read `vignette("tool_function")` before preparing the input data.

## Convert from species to characteristics
```{r conver2char}
# view input abundance
head(extract_summarized_experiment(processed_se)$abundance[, 1:5], 5)

# convert species abundance to characteristic abundance
char_abundance <- convert_sp2char(processed_se, transform='log10')

# view abundance after conversion
head(extract_summarized_experiment(char_abundance)$abundance[, 1:5], 5)
```

## Convert from species to ratio
```{r conver2ratio}
# view input abundance
head(extract_summarized_experiment(processed_se)$abundance[, 1:3], 5)

# convert species abundance to characteristic abundance
ratio_abundance <- convert_sp2ratio(processed_se, transform='log2')

# view abundance after conversion
head(extract_summarized_experiment(ratio_abundance )$abundance[, 1:3], 5)
```
